Efficient Planning for Sokoban with Answer Set Programming - A Case Study

### 2018-10-03 BATexFiles/Draft.tex
NEUE VERSION Draft.pdf ist viel weiter

\documentclass{report}

@@

\date{\today}

@@

@@

@@

\renewcommand*\contentsname{Summary}

@@

\paragraph{Complexity:} Gordon Wilfong published in 1991 that all motion planning problems with moving obstacles are NP-hard \cite{WilfongNPhard}. He proved this by reducing $3$-$Sat$ to such a problem. The question whether Sokoban is PSPACE-complete remained open until 1997, when Joseph C. Culberson published his paper "Sokoban is PSPACE-complete" \cite{PSpaceComplete}. He found a transformation algorithm to transform an LBA into a level of Sokoban, with the solution to the level having a number of steps in $\Theta (|w|+t(|w|))$, where $w$ is the input and $t(|w|)$ is the amount of transitions the LBA made.

@@

In my work I decided to research Sokoban planning with Answer Set Programming (ASP), a logic programming system designed to return answer sets to queries \cite{LifschitzASP}. I created different basic encodings and strategy implementations in a modular fashion. Those I will present to you in this chapter. Additionally I created a level portfolio, corresponding to different qualitative properties, e.g. level dimensions or number of boxes, and quantitative properties e.g. a seperation of the level into a zone where all the boxes are placed and a zone where all the targets are placed. The levels have two different types of structures. The first type is a rectangular level with the X and Y dimension differing by 1. The second type is created by connecting 2x2 fields at their edges, see \ref{fig:levelCrooked}.

@@

In the first few lines basic information is created, namely a $time/1$ predicate for each timestep (l.1), a $dir/2$ predicate for each of the four possible directions to move in (l.3) and an $at/4$ predicate to set the initial positions of objects as their position at time T=0.

@@

@@

\subsection{Split Encoding}
\lstinputlisting{../EncodingsClean/SokobanSplitEncoding.lp}
\begin{lstlisting}[caption={Split Encoding},label=list:SplitEnc]
\end{lstlisting}
The split encoding uses a technique common in ASP planning. When all coordinates are split into a vertical and a horizontal information the grounder does not have to ground all combinations of coordinates ($X * Y$) but only the sum of coordinates ($X + Y$). This has the advantage of less ground instances, but the disadvantage of a harder solving process. Just compare the ramifications in the plain encoding those in the split encoding. The plain encoding can deal with all ramifications in just three rules, while the split encoding needs ten rules. Obviously when splitting the coordinates into horizontal and vertical the rules deriving the new coordinates have to be twice as many. The other problem is that it is harder to find out whether two objects meet each other at a time. That is why the two frame axioms, signifying that boxes that are not touched do not move, have aggregates (ll.17-20). Those are used for simplifying complex rules; the first rule without the aggregates can be seen in listing \ref{list:BadFrame}.
\lstinputlisting{../EncodingsClean/BadSplitFrameAxiom.txt}
\begin{lstlisting}[caption={Frame Axiom X},label=list:BadFrame]
\end{lstlisting}
The last few changes are a second line for the initial readout of the level in lines 4 and 5, a few updates to the integrity constraints in lines 24 and 25 and a slightly different goal definition using the split coordinates.
%Here is another encoding, following a common ASP planning technique that aims to reduce ground instances. By splitting the horizontal and vertical coordinates, the program does not have to ground all combinations ($dimX * dimY$) of coordinates, but just each coordinate by itself (totaling $dimX + dimY$).% only grounds $(dimX + dimY) * horizon * N$ different positions instead of $dimX * dimY * horizon * N$.

@@

My third approach to deadlock detection (see listing \ref{list:Clutter}) recognizes a "clutter" of boxes (see Fig. \ref{fig:clutter}). This is a 2x2 square where each field is either a box or a wall, which would imply that all those boxes would have to stay in their places. In line 8 of the strategy implementation we can see that the box D does not have a target underneath, which qualifies the situation to be a deadlock. This type of deadlock happens a lot in crammed levels with a high box to field ratio. Those very common deadlocks are very important to find and remove to reduce the search tree in those levels efficiently.

\subsubsection{Move Evaluation}

@@

The last move evaluation seen in listing \ref{list:NoTake} tries to recognize moves that have no effect. This is done by checking for move instances $move(DX,DY,T)$ and $move(-DX,-DY,T-1)$ where no box was moved in timestep T-1. This would imply that the player moved back and forth without actually changing the gamestate, which means the plan is not relevant for finding the shortest plan in the level instance.
\chapter{Evaluation}
\chapter{Discussion}
compare chapter Related Work with chapter My Work \cite{LifschitzASP}

@@

>>>>

\documentclass[12pt,a4paper,oneside]{report}

@@

\date{30.09.2018}
\usepackage[utf8]{inputenc}
\usepackage[ngerman]{babel}
\usepackage[T1]{fontenc}

@@

\lstdefinelanguage{clingo}{
  keywordstyle=[1]\usefont{OT1}{cmtt}{m}{n},%
  keywordstyle=[2]\textbf,%
  keywordstyle=[3]\usefont{OT1}{cmtt}{m}{n},%\textit
  alsoletter={\#,\&},% 
keywords=[1]{not,from,import,def,if,else,return,while,break,and,or,for,in,del,and,class},% 
keywords=[2]{\#const,\#show,\#minimize,\#base,\#theory,\#count,\#external,\#program,\#script,\#end,\#heuristic,\#edge,\#project,\#show},%
  keywords=[3]{&,&dom,&sum,&diff,&show,&minimize},%
  morecomment=[l]{\#\ },%
  morecomment=[l]{\%\ },%
  commentstyle={\color{darkgray}}%
}
\lstset{numbers=left,numberblanklines=false,basicstyle=\ttfamily\footnotesize,language=clingo}
\usepackage{multirow}

@@

\usepackage[top=1.25in, bottom=1in, left=4cm, right=3.1cm]{geometry}

@@

\renewcommand*\contentsname{Table of Contents}

@@

\paragraph{Complexity:} Gordon Wilfong published in 1991 that all motion planning problems with moving obstacles are NP-hard \cite{WilfongNPhard}. He proved this by reducing $3$-$Sat$ to such a problem. The question whether Sokoban is PSPACE-complete remained open until 1997, when Joseph C. Culberson published his paper 'Sokoban is PSPACE-complete' \cite{PSpaceComplete}. He found a transformation algorithm to transform an LBA into a level of Sokoban, with the solution to the level having a number of steps in $\Theta (|w|+t(|w|))$, where $w$ is the input and $t(|w|)$ is the amount of transitions the LBA made.

@@

In my work I decided to research Sokoban planning with Answer Set Programming (ASP), a logic programming system designed to return answer sets to queries \cite{LifschitzASP}. I created different basic encodings and strategy implementations in a modular fashion. Those I will present to you in this chapter. Additionally I created a level portfolio, corresponding to different qualitative properties, e.g. level dimensions or number of boxes, and quantitative properties e.g. a seperation of the level into a zone where all the boxes are placed and a zone where all the targets are placed. The levels have two different types of structures. The first type is a rectangular level with the X and Y dimension differing by 1. The second type is created by connecting $2 \times 2$ fields at their edges, see \ref{fig:levelCrooked}.

@@

In the first few lines basic information is created, namely a $time/1$ predicate for each timestep (l.1), a $dir/2$ predicate for each of the four possible directions to move in (l.3) and an $at/4$ predicate to set the initial positions of objects as their position at time $T=0$.

@@

\\

@@

\\
\subsection{Split Encoding}
\lstinputlisting{../EncodingsClean/SokobanSplitEncoding.lp}
\begin{lstlisting}[caption={Split Encoding},label=list:SplitEnc]
The split encoding uses a technique common in ASP planning. When all coordinates are split into a vertical and a horizontal information the grounder does not have to ground all combinations of coordinates ($X \cdot Y$) but only the sum of coordinates ($X + Y$). This has the advantage of less ground instances, but the disadvantage of a harder solving process. Just compare the ramifications in the plain encoding those in the split encoding. The plain encoding can deal with all ramifications in just three rules, while the split encoding needs ten rules. Obviously when splitting the coordinates into horizontal and vertical the rules deriving the new coordinates have to be twice as many. The other problem is that it is harder to find out whether two objects meet each other at a time. That is why the two frame axioms, signifying that boxes that are not touched do not move, have aggregates (ll.17-20). Those are used for simplifying complex rules; the first rule without the aggregates can be seen in listing \ref{list:BadFrame}.
\lstinputlisting{../EncodingsClean/BadSplitFrameAxiom.txt}
\begin{lstlisting}[caption={Frame Axiom X},label=list:BadFrame]
The last few changes are a second line for the initial readout of the level in lines 4 and 5, a few updates to the integrity constraints in lines 24 and 25 and a slightly different goal definition using the split coordinates.
%Here is another encoding, following a common ASP planning technique that aims to reduce ground instances. By splitting the horizontal and vertical coordinates, the program does not have to ground all combinations ($dimX * dimY$) of coordinates, but just each coordinate by itself (totaling $dimX + dimY$).% only grounds $(dimX + dimY) * horizon * N$ different positions instead of $dimX * dimY * horizon * N$.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=0.2]{deadlocktypes}
\caption{'Clutter' in four configurations}
\label{fig:clutter}
\end{figure}

My third approach to deadlock detection (see listing \ref{list:Clutter}) recognizes a 'clutter' of boxes (see Fig. \ref{fig:clutter}). This is a $2 \times 2$ square where each field is either a box or a wall, which would imply that all those boxes would have to stay in their places. In line 8 of the strategy implementation we can see that the box D does not have a target underneath, which qualifies the situation to be a deadlock. This type of deadlock happens a lot in crammed levels with a high box to field ratio. Those very common deadlocks are very important to find and remove to reduce the search tree in those levels efficiently.
\\
\subsubsection{Move Evaluation}

@@

The last move evaluation seen in listing \ref{list:NoTake} tries to recognize moves that have no effect. This is done by checking for move instances $move(DX,DY,T)$ and $move(-DX,-DY,T-1)$ where no box was moved in timestep $T-1$. This would imply that the player moved back and forth without actually changing the gamestate, which means the plan is not relevant for finding the shortest plan in the level instance.
\chapter{Evaluation}
\begin{table}[]
\begin{tabular}{l|l|l|l|l|l|l|}
\cline{2-7}
                                  & \multicolumn{6}{l|}{Rectangular}                              \\ \cline{2-7} 
                                  & \multicolumn{3}{l|}{Small} & \multicolumn{3}{l|}{Large} \\ \hline
\multicolumn{1}{|l|}{\# of Boxes} & 4      & 5       & 6       & 6          & 8     & 10    \\ \hline
\multicolumn{1}{|l|}{Basic}       & 2.97   & 34.157  & 10.615  & 46.283     & 900   & 900   \\ \hline
\multicolumn{1}{|l|}{Clutter}     & 0.473  & 56.87   & 5.067   & 10.528     & 900   & 900   \\ \hline
\multicolumn{1}{|l|}{FF}          & 2.946  & 29.984  & 9.566   & 40.181     & 900   & 900   \\ \hline
\multicolumn{1}{|l|}{MOOB}        & 1.623  & 21.301  & 6.717   & 139.57     & 900   & 900   \\ \hline
\multicolumn{1}{|l|}{NTB}         & 1.051  & 57.824  & 14.575  & 151.681    & 900   & 900   \\ \hline
\multicolumn{1}{|l|}{WBTC}        & 2.655  & 30.85   & 9.611   & 42.185     & 900   & 900   \\ \hline
\multicolumn{1}{|l|}{Type1}       & 0.525  & 24.504  & 4.483   & 76.499     & 900   & 900   \\ \hline
\multicolumn{1}{|l|}{Type2}       & 0.523  & 24.728  & 4.483   & 78.336     & 900   & 900   \\ \hline
\end{tabular}
\end{table}
\begin{table}[]
\begin{tabular}{l|l|l|l|l|l|l|}
\cline{2-7}
                                  & \multicolumn{6}{l|}{Non-Rectangular}                             \\ \cline{2-7} 
                                  & \multicolumn{3}{l|}{Small} & \multicolumn{3}{l|}{Large} \\ \hline
\multicolumn{1}{|l|}{\# of Boxes} & 4       & 5       & 6      & 6       & 8       & 10     \\ \hline
\multicolumn{1}{|l|}{Basic}       & 2.648   & 8.132   & 2.655  & 900     & 900     & 900    \\ \hline
\multicolumn{1}{|l|}{Clutter}     & 1.1     & 3.107   & 2.127  & 900     & 900     & 900    \\ \hline
\multicolumn{1}{|l|}{FF}          & 2.234   & 8.054   & 2.473  & 900     & 900     & 900    \\ \hline
\multicolumn{1}{|l|}{MOOB}        & 2.23    & 7.953   & 2.466  & 900     & 900     & 900    \\ \hline
\multicolumn{1}{|l|}{NTB}         & 2.132   & 5.635   & 1.914  & 900     & 900     & 900    \\ \hline
\multicolumn{1}{|l|}{WBTC}        & 2.166   & 8.167   & 2.423  & 900     & 900     & 900    \\ \hline
\multicolumn{1}{|l|}{Type1}       & 1.496   & 4.736   & 2.285  & 900     & 900     & 900    \\ \hline
\multicolumn{1}{|l|}{Type2}       & 1.513   & 4.739   & 2.264  & 900     & 900     & 900    \\ \hline
\end{tabular}
\end{table}
\begin{table}[]
\begin{tabular}{l|l|l|l|l|l|}
\cline{2-6}
                              & \multicolumn{3}{l|}{Rectangular}         & \multicolumn{2}{l|}{Non-Rectangular} \\ \cline{2-6} 
                              & \multicolumn{2}{l|}{Small} & Large & \multicolumn{2}{l|}{Small}       \\ \hline
\multicolumn{1}{|l|}{\# of Boxes} & 5            & 8           & 10    & 4            & 5            \\ \hline
\multicolumn{1}{|l|}{Basic}   & 535.227      & 900         & 900   & 54.624       & 57.636       \\ \hline
\multicolumn{1}{|l|}{Clutter} & 193.626      & 900.031     & 900   & 26.104       & 26.351       \\ \hline
\multicolumn{1}{|l|}{FF}      & 586.862      & 900.007     & 900   & 51.58        & 52.374       \\ \hline
\multicolumn{1}{|l|}{MOOB}    & 305.945      & 900.013     & 900   & 51.586       & 52.889       \\ \hline
\multicolumn{1}{|l|}{NTB}     & 309.895      & 900.009     & 900   & 110.652      & 121.758      \\ \hline
\multicolumn{1}{|l|}{WBTC}    & 605.187      & 900.011     & 900   & 52.459       & 53.125       \\ \hline
\multicolumn{1}{|l|}{Type1}   & 38.671       & 534.863     & 900   & 35.324       & 35.413       \\ \hline
\multicolumn{1}{|l|}{Type2}   & 39.421       & 536.912     & 900   & 35.213       & 35.237       \\ \hline
\end{tabular}
\end{table}
\chapter{Discussion}
compare chapter Related Work with chapter My Work \cite{LifschitzASP}

@@

\newpage
\chapter*{Selbstständigkeitserklärung}
\newpage
\chapter*{Danksagung}
